Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays

### 2019-12-30 lit_review.Rmd
pdf for thesis

---
title: "Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian^[Science and Engineering Faculty, Queensland University of Technology.], Dianne Cook^[Faculty of Econometrics and Business Statistics, Monash University.] and Jessie Roberts^[Queensland University of Technology.]"
output:
  word_document:
    toc: no
    toc_depth: '3'
  pdf_document:
    dev: jpeg
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
geometry: margin = 1.2in
header-includes:
    - \usepackage{setspace}\doublespacing
link-citations: yes
csl: springer-vancouver.csl
fontsize: 11pt
bibliography: reviewbib.bib
yolo: yes
---

@@

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. It is easy to provide counts or incidence rates of the diagnoses of these areas.  This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

# 2. Traditional approaches for cancer map displays {#ch:choropleth_maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI, @BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

## 2.1 Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Figure 1: A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

Table: Table 1: <!--\label{tab:choropleths}--> A selection of choropleth cancer maps from online atlases.

@@

Table: Table 2: <!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

## 2.2 Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

@@

## 2.3 Limitations of choropleth displays {#ch:chorolimit}

@@

# 3. Contemporary alternatives to choropleth maps {#ch:alternatives}

## 3.1 Cartograms
<!-- 1496 words inc table-->

@@

```{r ggcartograms, fig.cap = "Figure 2: Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE, out.width='100%'}
knitr::include_graphics("figures/usa_grid.png")
```

Table: Table 3: <!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

### 3.1.1 Contiguous

@@

### 3.1.2 Non-contiguous

@@

### 3.1.3 Dorling

@@

## 3.2 Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [@NPR], [@FiveThirtyEight], [@WSJ], [@WP]. Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.

@@

## 3.3 Geofacet
<!-- 166 words -->
Hafen [@IGF] introduces the term geofacet to describe a grid display of small plots. The arrangement of tiles mimics the geographic topology. Geofaceting has the functionality that a statistical plot can be constructed in each facet for each geographic area. A tile map can communicate only one value per region in a visualization, while geofaceting is a more flexible visualization for communication as it increases the amount of information displayed. Virtually any type of plot can be shown in the tile, allowing displays of multiple variables or values per geographic entity. Creating the layout of a geofacet is manual, but once created can be used for any data on that geographic base.

```{r ggtilemap, fig.cap = "Figure 3: Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

## 3.4 Multivariate displays
<!-- 222 words -->

@@

# 4. Comparison and critique of alternative displays {#sec:compare}
<!-- 408 words -->

```{r auscartograms, fig.cap = "Figure 4: Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, out.width='100%', cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

## 4.1 Neither choropleth maps or cartograms perform well for Australia

@@

## 4.2 Limitations of alternative displays

@@

Table: Table 3: <!--\label{tab:methods}--> Summary of features and constraints of common mapping methods used to display cancer statistics (Y=Yes, N=No, S=Sometimes).

@@

# 5. User interaction {#ch:interacting}
<!-- 634 words -->

@@

<!-- ## 5.1 Under-utilized techniques-->

@@

```{r interacting, fig.cap = "Figure 5: Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for [Globocan 2018: Cancer Today ](http://globocan.iarc.fr/Pages/Map.aspx), (ii) Menus for variable selection and zooming on [Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), (iii) Menus for choosing variables and countries in [The Cancer Atlas ](http://canceratlas.cancer.org/data/#?view=map&metric=INCID_ALL_M),  (iv) Tabs for different indicators and cancer types in [Global Cancer Map](http://globalcancermap.com/), (v) Menus and toggles for variable and subset selection in [United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html).", fig.width = 12, fig.height = 9}
knitr::include_graphics("figures/interacting.png")
```

@@

# 6. Conclusions
<!-- 307 words -->

@@

<!--
## counts versus predictions

@SEMA
'Disease mapping is carried out to summarise spatial and spatio-temporal variation in risk.'
Descriptive statistic of the relationship between geography and a disease.
population, exposure, and health data may have point data or aggregated summaries (count data)
potentially be subject to a number of inaccuracies
diagnostic error or misclassification

Not only the counts, but population, migration, deaths and births are modelled across census years.

rapid visual summary of complex geographical information, descriptive purposes, to generate hypotheses, for surveillance, and to aid policy formation and resource allocation.

When using counts:
 small populations imply large variability in estimated rates

Decisions taken at this stage may be critical as apparent ‘clusters’ may depend crucially on the boundaries chosen in time or space: ‘The more narrowly the underlying population is defined, the less will be the number of expected cases, the greater will be the estimate of the excess rate, and often the more pronounced will be the statistical significance’

Using prediction:
assuming that neighboring μi are correlated leads to a ‘clustering’ prior

-->

<!-- presents forms of low level interactions with a user interface, these are the control menus for several online cancer atlases.
The user interface for these online atlases allow combinations of cancer type, gender, and other factors to be shown on the map display and the user to move between and/or compare different views of these variables. 
These controls may also change the geographies shown, such as continents or countries, states or counties. These displays can show familiar shapes for users to orient themselves on the map, then zoom in to explore densely populated communities.
Interactions with online atlases can also change the amount of detail shown.-->

<!-- Interactivity can be implemented within modern mapping methods to incorporate additional information and complexity without overloading the user.
Adding too much information to a map may create cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. 
Complexity and density of visualization methods can overwhelm novice decision makers however, experts are able to use the detail more readily when making-decisions [@cliburn2002design]. 
Supplementary information can be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop-down menus, data selection, zooming and panning.
Allowing users to explore the map for more information and providing flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication.
Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information [@maceachren1992visualizing, @van1994visualization]. For example, changing the population, age or other demographic variables.
The controls for basic interactive features are often placed outside of the plot space [@TGA], the map image is updated/replaced as the user interacts with the controls. 
Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection interactive capabilities.

Relationships between spatial areas and diseases can be explored with sophistication in nontraditional but still 'cognitively accessible' ways [@TNTEA]. This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution.

The interactive features of the publicly available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.

Interactivity and animation provide an alternative to the traditional approach of a single map for a single use, and enable the map to become more exploratory.
Interactivity allows users to drive their own exploration, controlling a display through variable controls, links, tooltips, and data selection tools [@TGA]. tooltips allow users to obtain additional information regarding specific geographic areas.

Development and adoption of this technology has allowed map users more control of the displayed message. 
-->

<!--to expand the communicative ability of a disease map. It is important to identify the primary questions that might be asked by the users of a map, and allow the display to change to best answer these questions. @CPISACA provide three questions that are generally asked when using a disease map:

- What is the mortality rate in a certain area?
- Are there geographic trends in the data, or regions of unusually high or low rates?
- Is the lung cancer mortality pattern similar to the pattern of smoking prevalence shown in a companion map?

A single map display may not effectively answer all three at the same time.
This is especially true when some areas are small and difficult to identify clearly in a single display or resolution. However, with the ability to interact with a map, or animate the display, map creators can specify and answer these questions by directing the attention of a map user to the areas of interest.
-->

>>>>

---
title: "Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian^[Science and Engineering Faculty, Queensland University of Technology.], Dianne Cook^[Faculty of Econometrics and Business Statistics, Monash University.] and Jessie Roberts^[Queensland University of Technology.]"
output:
  pdf_document:
    dev: jpeg
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: no
    toc_depth: '3'
geometry: margin = 1.2in
header-includes:
    - \usepackage{setspace}\doublespacing
link-citations: false
csl: springer-vancouver.csl
fontsize: 11pt
bibliography: reviewbib.bib
yolo: yes
---

@@

#  Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. It is easy to provide counts or incidence rates of the diagnoses of these areas.  This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

#  Traditional approaches for cancer map displays {#ch:choropleth_maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI, @BCM]. Faithful rendering of the geography, when combined with an appropriate color scheme, can reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. Choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

@@

##  Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, and geographic correlations [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. Cancer maps and atlases date back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "  A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atlases are described in Table 1.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

Table:  <!--\label{tab:choropleths}--> A selection of choropleth cancer maps from online atlases.

@@

Table:  <!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

##  Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

@@

##  Limitations of choropleth displays {#ch:chorolimit}

@@

#  Contemporary alternatives to choropleth maps {#ch:alternatives}

##  Cartograms
<!-- 1496 words inc table-->

@@

```{r ggcartograms, fig.cap = "  Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE, out.width='100%'}
knitr::include_graphics("figures/usa_grid.png")
```

Table:  <!--\label{tab:usa}--> Maps used to present statistics for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016.

@@

###  Contiguous

@@

###  Non-contiguous

@@

###  Dorling

@@

##  Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [@NPR], [@FiveThirtyEight], [@WSJ], [@WP]. Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.

@@

##  Geofacet
<!-- 166 words -->
Hafen [@IGF] introduces the term geofacet to describe a grid display of small plots. The arrangement of tiles mimics the geographic topology. Geofaceting has the functionality that a statistical plot can be constructed in each facet for each geographic area. A tile map can communicate only one value per region in a visualization, while geofaceting is a more flexible visualization for communication as it increases the amount of information displayed. Virtually any type of plot can be shown in the tile, allowing displays of multiple variables or values per geographic entity. Creating the layout of a geofacet is manual, but once created can be used for any data on that geographic base.

```{r ggtilemap, fig.cap = "  Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

##  Multivariate displays
<!-- 222 words -->

@@

#  Comparison and critique of alternative displays {#sec:compare}
<!-- 408 words -->

```{r auscartograms, fig.cap = "  Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, out.width='100%', cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

##  Neither choropleth maps or cartograms perform well for Australia

@@

##  Limitations of alternative displays

@@

Table:  <!--\label{tab:methods}--> Summary of features and constraints of common mapping methods used to display cancer statistics (Y=Yes, N=No, S=Sometimes).

@@

#  User interaction {#ch:interacting}
<!-- 634 words -->

@@

<!-- ##  Under-utilized techniques-->

@@

```{r interacting, fig.cap = "  Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for [Globocan 2018: Cancer Today ](http://globocan.iarc.fr/Pages/Map.aspx), (ii) Menus for variable selection and zooming on [Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), (iii) Menus for choosing variables and countries in [The Cancer Atlas ](http://canceratlas.cancer.org/data/#?view=map&metric=INCID_ALL_M),  (iv) Tabs for different indicators and cancer types in [Global Cancer Map](http://globalcancermap.com/), (v) Menus and toggles for variable and subset selection in [United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html).", fig.width = 12, fig.height = 9}
knitr::include_graphics("figures/interacting.png")
```

@@

#  Conclusions
<!-- 307 words -->

@@
